Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-03 response_to_reviewers_R1.Rmd

---
title: "Response to Reviewers"
date: "2023-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

Overall, the paper is now much more comprehensive, complete, and clear; I think it
will serve as a very useful introduction to the field of sports analytics. I have only a few
suggestions and points of confusion to share, which I hope the authors will examine before
submitting the final version. They are all fairly minor.

- If there is still space allowed in the abstract, I think you could consider explicitly
mentioning the three big ideas (expected value of a game state, win probability, and
measures of team strength). This would give readers a better idea of what is to come
and also allow people to find your article more easily if they are interested in one of
those specific topics.
- In Section 2.1.2, I am a little confused as to what Carroll et. al. did to extend the
previous model of Carter and Machol, since both efforts seem to involve some sort of
linearity assumption.
- Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
one didn’t know the definition before looking it up (and as it turns out the concept is
quite simple)
- In Section 2.1.3, you make the comment “they estimate that a better strategy would
be worth about 0.4 wins per season on average, a substantial amount in a season with
only 16 games.” This makes sense, but previously in Section 2.1.2 you mentioned
a potential difference of 3.8 wins per season, yet didn’t comment on the significant
magnitude of that effect. I’m not advocating for a particular change here, but the
authors could consider adding a small comment about the size of the effect Pelechrinis
et. al. estimated.
- At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
too conservative on 4th down decision making, but the effect may be overstated due
to the way in which to go yardage has been recorded in the past.” It’s unclear to me
whether the overstated effect is the one Lopez estimates in his paper, or if you are
referring to the estimated effects past studies which used less granular data.
- In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,
but it isn’t clear what those strategies are in reference to the previous discussion. Is
pulling the goalie optimal?
- In the second formula of Section 4.1, the right hand side of the equation has a 2, but
it should be an α.

> Thank you. This is fixed.

- I think Section 5.2 would benefit from a short introductory paragraph, perhaps which
introduces different types of sports betting, thus leading into the betting pools discus-
sion of Section 5.2.1.
- I notice you have several hyperlinks throughout–I’m not sure what the Wiley standards
are, but it might be helpful to distinguish links with a different color.

> Indeed. We await Wiley's instructions when we reach the typesetting phase.

## Minor comments from Reviewer 1

- Page 5, lines 33-34: ‘might reasonably’ is used twice in the same sentence

> We have changed the first "might reasonably" to "could"

- Page 6, line 23: ‘topics’ instead of ‘topic’

- Page 6, line 38: ‘go for it’ instead of ‘got for it’

- Page 6, lines 45-46: ‘coaches’ instead of ‘coached’ and ‘decision making’ instead of
‘decisions making’

- Page 7, line 7: ‘kickoffs’ instead of ‘kickoff’

- Page 8, line 51: Might be better to say “Earlier in Section 2” as opposed to just “In
Section 2”

- Page 11, line 56 and Page 12, line 5: ‘win-loss’ instead of ‘won-loss’

- Page 13, line 53: This is probably a matter of personal preference but “rating less than
2300” sounds more natural than “rating fewer than 2300”

- Page 14, line 35: ‘use’ instead of ‘uses’

- Page 17, line 27: I think the question in parentheses should be placed before the period
of the previous sentence

> We moved this.

- Page 18, line 40: ‘Lindsey’ instead of ‘Lindsay’

- Page 20, lines 13-15: Not sure why there are brackets around this sentence.

> We removed the brackets.

- Page 20, line 41: ‘Its’ not ‘It’s’

### 2023-01-03 sports_analytics.Rmd
address R1; #43

Second, the characterization of $S$ as having 24 states is only the simplest possible. 
The inning, or the score of the game, or even the weather, might reasonably be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes a more effective strategy.

@@

The concept of expected points in American football has many applications.  One of the most notable and well-studied topic is the evaluation of 4th down strategy.
There is near universal consensus in the literature that NFL teams have been too conservative in the past when making 4th down decisions.

@@

@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e. 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1. 
In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM]. However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making. 
With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data. 
Many past analyses of the decision to go for it or not on 4th down conclude that coached in the NFL are too conservative in their decisions making. 
@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.

@@

Another application of expected points is to evaluate kickoff decisions made by football coaches, as demonstrated by @urschel2011.  Specifically, they look at surprise on-sides kicks versus regular kickoff and the decision to accept a touchback versus returning the kickoff.  Using data from the 2009 NFL season, they conclude, as many have, that coaches in the NFL tend to make conservative decisions.

@@

In Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time. 
We then showed how this value could be used to analyze the relative effectiveness of certain strategies, with the simple idea that strategies that yield higher expected values are preferable. 
Generally, the goal of any sport is to score more points than the other team, which most often means trying to score as many points as possible, leading to a general strategy of maximizing expected points. 
However, there are situations in which maximizing the number of expected points is *not* the desired strategy.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
The methods in this Section apply equally well to multiplayer teams and single-player teams.

The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954, where the optimal value of $\alpha$ is 1.84.

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess.
Given two players $i$ and $j$ with unknown ratings $R_i$ and $R_j$, the probability $\Pi_{ij}$ of $i$ beating $j$ is defined as
$$
\Pi_{ij} = \frac{1}{1 + K^{(R_j - R_i)/400}} \,.
$$
In this formula, $K$ is commonly known as the $K$-factor, or development coefficient.
The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
For those that have never reached a rating of 2400 or better, $K=20$ is assigned. 
Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating fewer than 2300 [@fide2022rating].

@@

@koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.

@@

While @Lopez2018 uses betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
Neither paper finds strong evidence for inefficiencies in the markets. 
@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package). 
We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data. 
(How many of us can visualize PGN directly?)
In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.

@@

For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average. 
These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].

@@

Thus, these R packages enable research by making data more easily available. 
Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]

@@

As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics conferences have arisen.
The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
Many influential results have been showcased for the first time at these conferences. 
Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).

>>>>

Second, the characterization of $S$ as having 24 states is only the simplest possible. 
The inning, or the score of the game, or even the weather, could be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes a more effective strategy.

@@

The concept of expected points in American football has many applications.  One of the most notable and well-studied topics is the evaluation of 4th down strategy.
There is near universal consensus in the literature that NFL teams have been too conservative in the past when making 4th down decisions.

@@

@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "go for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e. 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1. 
In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM]. However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making. 
With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data. 
Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.

@@

Another application of expected points is to evaluate kickoff decisions made by football coaches, as demonstrated by @urschel2011. 
Specifically, they look at surprise on-sides kicks versus regular kickoffs and the decision to accept a touchback versus returning the kickoff.  
Using data from the 2009 NFL season, they conclude, as many have, that coaches in the NFL tend to make conservative decisions.

@@

Earlier in Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time. 
We then showed how this value could be used to analyze the relative effectiveness of certain strategies, with the simple idea that strategies that yield higher expected values are preferable. 
Generally, the goal of any sport is to score more points than the other team, which most often means trying to score as many points as possible, leading to a general strategy of maximizing expected points. 
However, there are situations in which maximizing the number of expected points is *not* the desired strategy.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on win-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
The methods in this Section apply equally well to multiplayer teams and single-player teams.

The impetus for all methods for estimating team strength is the realization that win-loss records are a noisy measure of team strength. 
As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^\alpha}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954, where the optimal value of $\alpha$ is 1.84.

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess.
Given two players $i$ and $j$ with unknown ratings $R_i$ and $R_j$, the probability $\Pi_{ij}$ of $i$ beating $j$ is defined as
$$
\Pi_{ij} = \frac{1}{1 + K^{(R_j - R_i)/400}} \,.
$$
In this formula, $K$ is commonly known as the $K$-factor, or development coefficient.
The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating less than than 2300 [@fide2022rating].

@@

@koopman2015dynamic use a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.

@@

While @Lopez2018 use betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
Neither paper finds strong evidence for inefficiencies in the markets. 
@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package). 
We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data
(How many of us can visualize PGN directly?).
In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.

@@

For example, before George Lindsey's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average. 
These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].

@@

Thus, these R packages enable research by making data more easily available. 
Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).

@@

As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics conferences have arisen.
The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
Its West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
Many influential results have been showcased for the first time at these conferences. 
Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).
